---
title: "hw2"
author: "Lai Wei"
format: 
  html:
    embed-resources: true
editor: visual
---

The link to the GitHub repository of this assignment is: https://github.com/Lwei493/Homework2

# Problem 1: Random Walk with Momentum

## a.Simulation function:

### i. Write a function (momentum_walk(n, p)) that simulates this process and returns the particle’s final position.

### ii. Sanitize the input of your function: 𝑛≥1 , 0≤𝑝≤1

```{r}
#' momentum_walk
#'
#' @param n the number of steps, non-negative integer (n >= 1)
#' @param p the probability of preserving the previous direction, single numeric value between 0 and 1 
#'        (0≤𝑝≤1)
#' @return numeric vector containing the final x and y coordinates
#' 
#' @examples
#' momentum_walk(10, 0.6)
#' momentum_walk(100, 0.8)
#' 
momentum_walk <- function(n, p) {
  if (n < 1 | n %% 1 != 0 | length(n) != 1 | !is.numeric(n) | is.na(n)) {
    stop("n must be a non-negative integer.")
  }
  if (p < 0 | p > 1 | length(p) != 1 | !is.numeric(p) | is.na(p)) {
    stop("p must be a single numeric value between 0 and 1.")
  }
  
  # original coordinate
  x <- 0
  y <- 0
  
  dx <- c(1, -1, 0, 0) # right and left
  dy <- c(0, 0, 1, -1) # up and down
  
  # first step, choose one of the four directions with equal probability
  previous_step <- sample(1:4, 1) 
  x <- x + dx[previous_step]
  y <- y + dy[previous_step]

  if (n == 1) return(c(x,y))
  
  if (n > 1) {
    for (i in 1:n-1) {
      if (runif(1) <= p) {
        # p, moves in the same direction as the previous step
        current_step <- previous_step 
      } else {
        # 1-p, chooses one of the four directions at random
        current_step <- sample(1:4, 1) 
      }
      
      x <- x + dx[current_step]
      y <- y + dy[current_step]
      
      previous_step <- current_step
    }
  }
  
  return(c(x, y))
}

# Example
momentum_walk(10, 0.6)
momentum_walk(100, 0.8)
# Because the process is random, the results will generally differ each time.
```

## b. Implement the simulation in two different ways:

### Version 1: write the function momentum_walk1(n, p) using a loop ( a for loop or a while loop).

```{r}
#| echo: true
#' momentum_walk1 (using a loop)
#'
#' @param n the number of steps, non-negative integer (n >= 1)
#' @param p the probability of preserving the previous direction, single numeric value between 0 and 1 
#'        (0≤𝑝≤1)
#' @return numeric vector containing the final x and y coordinates
#' 
#' @examples
#' momentum_walk1(10, 0.6)
#' momentum_walk1(100, 0.8)
#' 
momentum_walk1 <- function(n, p) {
  if (n < 1 | n %% 1 != 0 | length(n) != 1 | !is.numeric(n) | is.na(n)) {
    stop("n must be a non-negative integer.")
  }
  if (p < 0 | p > 1 | length(p) != 1 | !is.numeric(p) | is.na(p)) {
    stop("p must be a single numeric value between 0 and 1.")
  }
  
  # original coordinate
  x <- 0
  y <- 0
  
  dx <- c(1, -1, 0, 0) # right and left
  dy <- c(0, 0, 1, -1) # up and down
  
  # first step, choose one of the four directions with equal probability
  previous_step <- sample(1:4, 1) 
  x <- x + dx[previous_step]
  y <- y + dy[previous_step]

  if (n == 1) return(c(x,y))
  
  if (n > 1) {
    for (i in 1:n-1) {
      if (runif(1) <= p) {
        # p, moves in the same direction as the previous step
        current_step <- previous_step 
      } 
      else {
        # 1-p, chooses one of the four directions at random
        current_step <- sample(1:4, 1) 
      }
      
      x <- x + dx[current_step]
      y <- y + dy[current_step]
      
      previous_step <- current_step
    }
  }
  
  return(c(x, y))
}

# Example
momentum_walk1(10, 0.6)
momentum_walk1(100, 0.8)
```

### Briefly explain the logic.

Version 1 uses a for loop to perform the simulation step-by-step.

The function iterates from step 2 to step n, with the length n-1.

In each iteration, it generates a random number to decide whether to keep the previous direction (with probability p) or randomly select a new direction (with probability 1-p).

It then updates the x and y coordinates and the previous_step variable.

This imperative approach clearly reflects the sequential nature of the random walk, but it may be less efficient in R for very large n.

### Version 2: write the function momentum_walk2(n, p) using one of the apply functions (choose an appropriate member of the apply family).

```{r}
#| echo: true
#' momentum_walk2 (using one of the apply functions)
#'
#' @param n the number of steps, non-negative integer (n >= 1)
#' @param p the probability of preserving the previous direction, single numeric value between 0 and 1 
#'        (0≤𝑝≤1)
#' @return numeric vector containing the final x and y coordinates
#' 
#' @examples
#' momentum_walk2(10, 0.6)
#' momentum_walk2(100, 0.8)
#' 
momentum_walk2 <- function(n, p) {
  if (n < 1 | n %% 1 != 0 | length(n) != 1 | !is.numeric(n) | is.na(n)) {
    stop("n must be a non-negative integer.")
  }
  if (p < 0 | p > 1 | length(p) != 1 | !is.numeric(p) | is.na(p)) {
    stop("p must be a single numeric value between 0 and 1.")
  }
  
  # original coordinate
  x <- 0
  y <- 0
  
  dx <- c(1, -1, 0, 0) # right and left
  dy <- c(0, 0, 1, -1) # up and down
  
  # first step, choose one of the four directions with equal probability
  previous_step <- sample(1:4, 1) 
  x <- x + dx[previous_step]
  y <- y + dy[previous_step]
  
  if (n == 1) return(c(x,y)) 
  
    if (n > 1) {
    invisible(lapply(1:(n - 1), function(i) {
      if (runif(1) <= p) {
        current_step <- previous_step
      } 
      else {
        current_step <- sample(1:4, 1)
      }
      
      x <<- x + dx[current_step]
      y <<- y + dy[current_step]
      
      previous_step <<- current_step
    }))
    }
  return(c(x, y))
}

# Example
momentum_walk2(10, 0.6)
momentum_walk2(100, 0.8)
```

### Briefly explain the logic.

Version 2 uses lapply from the apply family to replace the loop.

It applies an anonymous function to a sequence of length n-1.

Inside the anonymous function, the same logic is applied to determine the direction and update the coordinates.

Because lapply creates a separate local scope for each iteration, the super-assignment operator (\<\<-) is used to modify the state variables (x, y, and previous_step) in the global environment.

Finally, invisible( ) is wrapped around the 'lapply' to prevent the console from printing a large list of NULL values.

## c. Reproducibility

### Random simulations normally give different results each time they are run. Find a way to control R’s random-number generation so that your results are reproducible.

### Choose version 1 in part 1-b as the implementation for this question.

```{r}
#| echo: true
# n = 10, p = 0.6

# First time
set.seed(17)  
first_10 <- momentum_walk1(10, 0.6)
print(first_10)

# Second time. Check whether the same result can be reproduced
set.seed(17)
second_10 <- momentum_walk1(10, 0.6)
print(second_10)

cat("Are the results are reproducible for n = 10?", 
    identical(first_10, second_10), "\n\n")


# n = 1000, p = 0.6

# First time
set.seed(17)
first_1000 <- momentum_walk1(1000, 0.6)
print(first_1000)

# Second time. Check whether the same result can be reproduced
set.seed(17)
second_1000 <- momentum_walk1(1000, 0.6)
print(second_1000)

cat("Are the results identical for n = 1000?", 
    identical(first_1000, second_1000), "\n")
```

### Explain what you did and why it works.

### What I did:

I used the set.seed( ) function before each simulation run. I chose a fixed integer (e.g., 17) as the seed.

To demonstrate reproducibility, I ran the momentum_walk1 function twice for n=10, p=0.6 and n=1000, p=0.6, resetting the seed to the same value before each run.

### Why it works:

Computers generate "random" numbers using deterministic algorithms called pseudo-random number generators (PRNGs).

The sequence of numbers they produce is determined entirely by an initial value called the seed.

By default, R generates a seed from the system clock, which is why results differ each time when running a simulation.

By explicitly setting the seed with set.seed( ), we can force the PRNG to start from the same state, ensuring that the exact same sequence of random numbers is generated.

Therefore, as long as the seed and the code are identical, the simulation outputs will be exactly reproducible.

## d. Does momentum matter?

### i. For each value of p, simulate a walk of 1000 steps.

### Consider three different momentum probabilities: p = 0; p = 0.5; p = 0.9.

### For a final location (x,y), define the distance from the starting point

### Choose version 1 in part 1-b as the implementation for this question.

```{r}
#| echo: true
set.seed(17) # ensure a fair comparison between the three different momentum probabilities
coord1 <- momentum_walk1(1000, 0)
coord2 <- momentum_walk1(1000, 0.5)
coord3 <- momentum_walk1(1000, 0.9)
distance <- function(coord) {
  sqrt(coord[1]^2 + coord[2]^2)
}
D1_0 <- distance(coord1)
D2_0.5 <- distance(coord2)
D3_0.9 <- distance(coord3)
cat("For p = 0, the final coordinate is", coord1, "and distance from the starting point =", round(D1_0, 4), "\n", "For p = 0.5, the final coordinate is", coord2, "and distance from the starting point =", round(D2_0.5, 4), "\n", "For p = 0.9, the final coordinate is", coord3, "and distance from the starting point =", round(D3_0.9, 4))
```

### ii. Does increasing momentum appear to increase or decrease the final distance?

To ensure a fair comparison, I set the same random seed before running each simulation. This ensures the only changing factor is the momentum probability.

As the momentum probability(p) increases from 0 to 0.9, the final position of the particle tends to move farther and farther away from the starting point.

Increasing momentum appears to increase the final distance.

### iii. Explain your reasoning (optional).

When p=0, It is a standard random walk, each step is completely independent. The particle chooses a random direction at every step. Because the direction changes frequently, the movements often cancel each other out (e.g., one step right and one step left). Mathematically, the expected squared distance grows only linearly with the number of steps, resulting in a relatively small net displacement after 1000 steps.

When p \> 0, the particle has a tendency to continue moving in its current direction. This reduces the frequency of direction reversals. Since particles may move again in a similar direction, the movement can accumulate in one direction, resulting in a greater net displacement. Therefore, higher momentum leads to a larger final distance from the origin.

## e. Monte Carlo simulation

### i. For each of: p = 0, 0.25, 0.5, 0.75, 0.9, simulate 1,000 independent walks, each consisting of 1,000 steps.

### ii. For each value of p, calculate the mean final distance

### iii. Print out your results in a table with two columns: p and mean_distance.

### Choose version 1 in part 1-b as the implementation for this question.

```{r}
#| echo: true
p_value <- c(0, 0.25, 0.5, 0.75, 0.9)

set.seed(17)

mean_distances <- function(p) {
  distances <- replicate(1000, {
    coord <- momentum_walk1(1000, p)
    sqrt(coord[1]^2 + coord[2]^2)  
  })
  return(mean(distances))  
}

mean_distances <- vapply(p_value, mean_distances, FUN.VALUE = numeric(1))

results_table <- data.frame(p = p_value, mean_distances = round(mean_distances, 4))

print(results_table)
```

### iv. How does the mean final distance change as p increases?

As p increases, the mean final distance increases significantly.

## f.Compare computational performance

### Use the microbenchmark package to compare the speed of your two implementations. Compare their performance when n = 1000, p = 0.6, and n = 100000, p = 0.6.

```{r}
#| echo: true
library(microbenchmark)

set.seed(17)

# n = 1000, p = 0.6
microbenchmark(
  Loop  = momentum_walk1(1000, 0.6),
  Apply = momentum_walk2(1000, 0.6)
)

# n = 100000, p = 0.6
microbenchmark(
  Loop  = momentum_walk1(100000, 0.6),
  Apply = momentum_walk2(100000, 0.6)
)

```

### i. Which implementation is faster?

momentum_walk1 (the ”for“ loop implementation) is faster than momentum_walk2 (the ”lapply“ implementation).

### ii. Does the difference become larger for the larger input?

Yes, the difference becomes larger for the larger input.

When n=1000, the mean execution time for the for loop is 2.644582 milliseconds, for lapply is 4.707638.

When n=100000, the mean execution time for the for loop is 282.3917 milliseconds, for lapply is 413.0722.

The performance gap scales up significantly as the number of steps increases.

### iii. Why might the implementations have different execution times?

The primary reason is how R manages environments and variable assignments in loops versus apply family functions.

In momentum_walk1 (the ”for“ loop), the variables are updated directly in the function's local environment. This is a low-overhead operation.

However, in momentum_walk2 (”lapply“), each iteration creates a new environment. Because the anonymous function needs to update variables outside its own scope, it must use the super-assignment operator (\<\<-). This forces R to search up the environment chain to find and modify the parent variables, which is computationally expensive. Additionally, lapply allocates memory for a return list during the iterations. These environment lookups and memory allocations add significant overhead, making the apply version slower, especially for large n.

# Problem 2: Monte Carlo Simulation of Campus Bus Arrivals

## Use a Monte Carlo simulation to estimate the average total number of passengers who board the bus during one day, under five assumptions:

## Simulate 10,000 days and use your simulation results to estimate the expected total number of passengers who board the bus per day.

```{r}
#| echo: true
one_day <- function() {
  # 6 AM to 8 AM (3 hours): The number of passengers per hour follows a Poisson distribution with mean 4.
  period1 <- sum(rpois(3, lambda = 4))
  
  # 9 AM (1 hour): a Normal distribution with mean 45 and variance 16.
  period2 <- rnorm(1, mean = 45, sd = sqrt(16))
  
  # 10 AM to 3 PM (6 hours): a Poisson distribution with mean 15.
  period3 <- sum(rpois(6, lambda = 15))
  
  # 4 PM (1 hour): a Normal distribution with mean 55 and variance 25.
  period4 <- rnorm(1, mean = 55, sd = sqrt(25))
  
  # 5 PM to 10 PM (6 hours): a Poisson distribution with mean 9.
  period5 <- sum(rpois(6, lambda = 9))
  
  # total number of passengers who board the bus during one day
  return(period1 + period2 + period3 + period4 + period5)
}

# Simulate 10,000 days
days_totals <- replicate(10000, one_day())

expected_passengers <- mean(days_totals)

cat("The expected total number of passengers who board the bus per day is:", 
    round(expected_passengers, 0))
```

# Problem 3: Linear Regression

## a. Data preparation and de-identification

### Download the YouTube Super Bowl commercials data

```{r}
youtube <- read.csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-03-02/youtube.csv"
)
```

### Remove all variables that could uniquely identify a particular commercial. This includes variables such as the brand, title, URLs, YouTube channel, and publication information.

### After removing these variables, report the dimensions of the resulting dataset.

```{r}
#| echo: true
variables <- names(youtube)

variables_remove <- c(
  "brand", 
  "superbowl_ads_dot_com_url", 
  "youtube_url", 
  "id", 
  "kind",
  "etag", 
  "published_at", 
  "title", 
  "description", 
  "thumbnail", 
  "channel_title",
  "category_id"
)

youtube1 <- youtube[, !is.element(variables, variables_remove)]

dimension <- dim(youtube1)

cat("The dimensions of the resulting dataset(youtube1) are", dimension[1], "rows and", dimension[2], "columns.")
```

## b. Exploring potential outcomes

### Consider each of the following YouTube engagement variables.

```{r}
#| echo: true
vars <- c("view_count", "like_count", "dislike_count", "favorite_count", "comment_count")

# Plot histograms to examine the distribution
par(mfrow = c(2, 3))  
for (var in vars) {
  hist(youtube1[[var]], main = var, xlab = var, col = "blue")
}
par(mfrow = c(1, 1))

for (var in vars) {
  mean <- mean(youtube1[[var]], na.rm = TRUE)
  median <- median(youtube1[[var]], na.rm = TRUE)
  cat(var, ": Mean =", mean, ", Median =", median, "\n")
}
```

### Decision for Each Variable:

View counts: 2. should be transformed.\
The distribution is heavily right-skewed, the mean (1,407,556) is vastly larger than the median (41,379). This is due to videos with millions of views.

Like counts: 2. should be transformed.\
Highly right-skewed. The mean (4146.031) is vastly larger than the median (130). Because some videos have relatively very high numbers of likes

Dislike counts: 2. should be transformed.\
Highly right-skewed, with many videos having zero or very few dislikes. The mean (833.5378) is vastly larger than the median (7).

Favorite counts: 3. not appropriate.\
The mean and median are both 0. The favorite_count variable consists entirely of zeros or missing values (NA).

```{r}
#| echo: true
cat("Summary of favorite_count:\n")
summary(youtube1$favorite_count)
cat("Number of Zeros:", sum(youtube1$favorite_count == 0, na.rm = TRUE), "\n")
cat("Number of NAs:", sum(is.na(youtube1$favorite_count)), "\n")
```

This may be because YouTube has disabled the "favorite" feature. As a result, the variance of this column is zero, making it unsuitable to use as the dependent variable in a linear regression model. Moreover, even after transformation, the result would still be zero. Therefore, this variable is not appropriate to use as the outcome in a linear regression model.

Comment counts: 2. should be transformed.\
Heavily right-skewed. The mean (188.6396) is vastly larger than the median (10).

### Perform an appropriate transformation and use the transformed variable for the remainder of the problem.

```{r}
vars_transform <- c("view_count", "like_count", "dislike_count", "comment_count")

for (var in vars_transform) {
  new_var <- paste0("trans_", var)
  youtube1[[new_var]] <- log(youtube1[[var]] + 1)
}

summary(youtube1[, paste0("trans_", vars_transform)])

# Plot histograms to examine the distribution after transformation
new_vars <- paste0("trans_", vars_transform)
par(mfrow = c(2, 3))  
for (var in new_vars) {
  hist(youtube1[[var]], main = var, xlab = var, col = "green")
}
par(mfrow = c(1, 1))
```

## c. Fitting linear regression models

### For each outcome from part 3-b that you determined was appropriate for linear regression, fit separate linear regression models to investigate its relationship with each of the seven binary characteristics of the commercials.

### Use one binary commercial characteristic as the primary predictor. Include year as a continuous covariate. Report the estimated coefficient for the commercial characteristic and its p-value.

```{r}
#| echo: true
outcomes <- c("trans_view_count", "trans_like_count", "trans_dislike_count", "trans_comment_count")

predictors <- c("funny", "show_product_quickly", "patriotic", "celebrity", "danger", "animals", "use_sex")

# change the binary characteristics to 0/1
youtube1[predictors] <- lapply(youtube1[predictors], as.numeric)

results <- data.frame()

for (outcome in outcomes) {
  for (predictor in predictors) {
    formula_3c <- paste(outcome, "~", predictor, "+ year")

    model <- lm(as.formula(formula_3c), data = youtube1)
    
    coef <- summary(model)$coefficients[predictor, "Estimate"]
    p <- summary(model)$coefficients[predictor, "Pr(>|t|)"]
    
    results <- rbind(results, data.frame(
      Outcome = outcome,
      Predictor = predictor,
      Estimate = coef,
      P_Value = p
    ))
  }
}

print(results)
```

### 1. Which commercial characteristics have a statistically significant association with each engagement outcome at the 𝛼 =0.05 significance level?

```{r}
#| echo: true
significant_results <- results[results$P_Value < 0.05, ]
cat("\nThe commercial characteristics have a statistically significant association with each engagement outcome at the 𝛼 =0.05 significance level are", nrow(significant_results))
```

### 2. For each statistically significant association, state whether the association is positive or negative.

Based on 3-c-1, there is no significant association.

### 3. Identify which commercial characteristic has the largest positive estimated association with view counts.

```{r}
#| echo: true
view_results <- results[results$Outcome == "trans_view_count" & results$Estimate > 0, ]
best <- view_results[which.max(view_results$Estimate), ]
cat("\nThe commercial characteristic which has the largest positive estimated association is:", best$Predictor, "(estimated coefficient =", best$Estimate, ")\n")
```

## d. Estimating regression coefficients using matrix algebra

### Choose the binary commercial characteristic that had the smallest p-value for predicting view counts in part 3-c.

```{r}
#| echo: true
smallest_p <- view_results[which.min(view_results$P_Value), ]

corresponding_var <- smallest_p$Predictor

cat("The binary characteristic with the smallest p-value for view counts is:", 
    corresponding_var, "(p =", smallest_p$P_Value, ")\n")
```

### Construct the appropriate design matrix X, including the intercept. Without using lm(), calculate the estimated regression coefficients using the matrix formula.

```{r}
youtube1_cleanNA <- youtube1[!is.na(youtube1$trans_view_count) &
  !is.na(youtube1[[corresponding_var]]) &
  !is.na(youtube1$year),
]
y <- youtube1_cleanNA$trans_view_count
n <- length(y)

X <- cbind(
  Intercept = rep(1, n),
  beta1 = youtube1_cleanNA[[corresponding_var]],  
  beta2 = youtube1_cleanNA$year
)

beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y

cat("\nCoefficients calculated via matrix algebra:\n")
print(beta_hat)
```

### Fit the same model using lm() and confirm that the estimated coefficients are the same, up to numerical precision.

```{r}
formula_3d <- paste("trans_view_count ~", corresponding_var, "+ year")

lm_model <- lm(as.formula(formula_3d), data = youtube1)

cat("\nEstimated coefficients calculated via lm():\n")
print(coef(lm_model))

cat("\nDifference between the two methods:\n")
print(beta_hat - coef(lm_model))
all.equal(
  as.vector(beta_hat),
  as.vector(coef(lm_model))
)
```

The coefficients calculated via the matrix formula are identical to those from lm() up to numerical precision.\
The maximum absolute difference is approximately 1.05 \times 10\^-10 (for the intercept) , which is negligible and attributable to floating-point rounding errors.
